---
title: ""
author: "Giorgio Arcara"
date: ""
#format: revealjs
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
---

<!-- the section below is shown only if rendering is beamer from latex. (and format:revealjs is commmented) -->

::: {.content-visible when-format="beamer"}
```{=latex}
% TITLE SLIDES
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{2.1 - Classical Test Theory}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{

\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```
:::

## Classical Test Theory

A first teory on tests
\vfill
```{=latex}
\url{https://en.wikipedia.org/wiki/Classical_test_theory}
```

## The cardinal principle of CTT is the following formula:
```{=latex}
\begin{center}
$\onslide<2->{X = }T \onslide<3->{+ E}$
\end{center}

$T$ = true score

\onslide<2->{$X$ = observed score}

\onslide<3->{$E$ = error}
\vfill
\onslide<4->True Score and Error are independent on each other

```

## An example
```{=latex}
$T$ = 10

$E$ = 2 
\pause

$X$ = 12
```

## An example
```{=latex}
$T$ = 10

$E$ = -3
\pause

$X$ = 7
```

## Another example
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4

set.seed(30)
plot(0,0, xlim=c(0,20), ylim=c(0,10), col=NULL, main="Distribution of observed scores", xlab="")
abline(v=12, lty=2, lwd=2)
legend("topleft", lty=2, lwd=2, legend="True Score")
```

## Another example
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4

#### CREATE PLOT FUNCTION
obs_plot = function(n, my_seed, my_sd){
  plot(0,0, xlim=c(0,20), ylim=c(0,10), col=NULL, main="Distribution of observed scores", xlab="")
abline(v=12, lty=2, lwd=2)
legend("topleft", lty=c(2,1), lwd=2, legend=c("True Score", "Observed Score"), col=c("black", "gray40"))
set.seed(my_seed)
err = rnorm(n, mean=0, sd=my_sd)
if (n>0){
segments(x0=c(12)+err, y0=-1, x1=(12)+err, y1=1.5, lwd=2, col="gray40")
}
}

obs_plot(n=0, my_seed=20, my_sd=3)

```
## Another example
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4

#### CREATE PLOT FUNCTION
obs_plot = function(n, my_seed, my_sd){
  plot(0,0, xlim=c(0,20), ylim=c(0,10), col=NULL, main="Distribution of observed scores", xlab="")
abline(v=12, lty=2, lwd=2)
legend("topleft", lty=c(2,1), lwd=2, legend=c("True Score", "Observed Score"), col=c("black", "gray40"))
set.seed(my_seed)
err = rnorm(n, mean=0, sd=my_sd)
if (n>0){
segments(x0=c(12)+err, y0=-1, x1=(12)+err, y1=1.5, lwd=2, col="gray40")
}
}

obs_plot(n=1, my_seed=30, my_sd=3)

```

## Another example
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4


obs_plot(n=3, my_seed=20, my_sd=3)

```

## Another example
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 14
#| fig-height: 6
par(mfrow=c(1,2))
obs_plot(n=100, my_seed=20, my_sd=3)
obs_plot(n=100, my_seed=20, my_sd=1)

```



## The usual situtation
```{=latex}
$X$ = 9
\pause

$T$ = ?
```



## Two main aims of Classical Test Theory

- having tests with lowest possible error
- estimating the error

## Concepts of Classical Test Theory

Two relevant concepts of Classical Test Theory are:
```{=latex}
\vfill
\begin{itemize}
\item \textbf{Validity} \pause (Are we measuring what we intend to measure?) \pause \\
\item \textbf{Reliability} \pause (How precise or consistent is our measurement?)
\end{itemize}
\pause
\vfill
\emph{See upcoming lessons for more details}
```

## Limits of Classical Test Theory

- Test ability and item difficulty are not modeled separately \pause
- Is error really indepentent from true score?

## Is error really independent from true score?
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4

plot(0,0, xlim=c(0,20), ylim=c(0,10), col=NULL, main="Distribution of observed scores", xlab="")
abline(v=20, lty=2, lwd=2)
legend("topleft", lty=2, lwd=2, legend="True Score")
```


