---
title: ""
author: "Giorgio Arcara"
date: ""
#format: revealjs
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
---

<!-- the section below is shown only if rendering is beamer from latex. (and format:revealjs is commmented) -->

::: {.content-visible when-format="beamer"}
```{=latex}
% TITLE SLIDES
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{5.1 - Assessing Validity: Correlations}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{

\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## Correlation

Correlation is one of the most important statistical analyses you may encounter in psychological testing (and not only).


## Correlation

It is a number that expresses the relation between *two* numerical variables.

## Correlation Values

It ranges from -1 to 1.

- A correlation value of 1 indicates perfect positive correlation
- A correlation value of -1 indicates perfect negative correlation
- A correlation value of 0 indicates no correlation between the variables.

## A scatterplot
```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=0.6, method="none")


```

## Examples - 1

```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=0.6)


```
## Examples - 2

```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=-0.6)

```
## Examples - 3

```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=0.99)

```
## Examples - 4
```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=-0.99)

```


## Can you guess the correlation?
```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(42)
plot_sim_corr(n=200, rho=0, plot_title = F)

```

## Types of correlation

So far I have shown results for **Pearson's correlation** (indicated with **r**)

There are many types of correlations; the most used in basic Psychometrics are Pearson's correlation and **Spearman's correlation**, indicated with **$\rho$**

## Spearman correlation

Spearman correlation should be used with ordinal data or when the assumption of normality is violated.

More simply, it is also more appropriate when outliers are present, as it is considered "robust" against them.

## When to use which?

| Condition | Use |
|-----------|-----|
| Continuous, normally distributed data | Pearson (*r*) |
| Ordinal data | Spearman (*$\rho$*) |
| Non-normal distribution or outliers present | Spearman (*$\rho$*) |

## Example of types of correlations

```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=4, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(45)
plot_sim_corr(n=200, rho=0.6, method="both")

```
## Adding an outlier

```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=3.5, fig.align='center', out.width="70%"}
source("R_functions/sim_corr.R")

set.seed(45)
outx1=15
outx2=0
plot_sim_corr(n=200, rho=0.6, addx1 = outx1, addx2 = outx2, method="both")
points(x=outx1, y=outx2, pch=19, col="black")

```
\footnotesize Note that the data are the same as in the plot in the slide before, except for the black point.

## Note!

When interpreting correlations, _scatterplots_ (i.e. the graphs used to represent relationships between two variables) are very important, but rarely reported.


## Correlation in Psychometrics

Correlations are used in many important aspects of Psychometrics; we encountered them for Convergent/Discriminant validity, but also for ecological validity.

We will encounter them also for _Reliability_ (next week).

## Interpreting correlations

Here is a common classification of correlation strength.

| r or R | r or R | Strength of association |
|--------|--------|------------------------|
| +0.8 to +1.0 | -1.0 to -0.8 | Perfect or very strong association |
| +0.6 to +0.8 | -0.8 to -0.6 | Strong association |
| +0.4 to +0.6 | -0.6 to -0.4 | Moderate association |
| +0.2 to +0.4 | -0.4 to -0.2 | Weak association |
| +0.0 to +0.2 | -0.2 to 0.0 | Very weak or no association |


## What is the "right" correlation value in the case of Validity?

There is no straightforward way to interpret whether a specific correlation is adequate or not in the context of validity.
\pause

For example, even a weak correlation can be adequate in the context of ecological validity.
\pause

**Read carefully the article and check whether the values are reasonably motivated by the authors.**

## Comparing correlation values
It is easier to interpret correlations for validity when discriminant validity is present.

In this case you can check whether the values for convergent validity are higher than those for discriminant validity.
\pause

## An example of interpreting validity

E.g. suppose you are checking a test that is supposed to measure memory problems in daily life.

Suppose you find that the correlation with an external measure of memory is r = 0.4.

\pause
The authors claim that this is proof of validity, as there is no expectation of very strong correlations.

## Two potential scenarios
_SCENARIO 1_

The authors also present discriminant validity and the correlation with a depression scale is r = 0.1.
\pause

_SCENARIO 2_

The authors also present discriminant validity and the correlation with a depression scale is r = 0.5.
\pause


## Correlation Matrix
```{r, echo=FALSE, output=TRUE, fig.width=4,fig.height=3.5, fig.align='center', out.width="70%"}
library("corrplot")

mat = matrix(c(1, -0.7, -0.6, -0.01,
               -0.7, 1, 0.5, 0.03,
               -0.6, 0.5, 1, 0.02,
               -0.01, 0.03, 0.02, 1
               ), byrow=TRUE, ncol=4)

colnames(mat)=rownames(mat)=c("Anxiety Scale", "Well-Being Scale", "Quality of Life Scale", "Age")


# isSymmetric.matrix(mat)
corrplot(mat, method = "number", col = "black", cl.pos = "n" )


```

## Correlation Matrix (another visualization)
```{r, echo=FALSE, output=TRUE, fig.width=4, fig.height=3.5, fig.align='center', out.width="80%"}

corrplot(mat, method = "ellipse", cl.offset = 5, cl.ratio=0.5)

```
