---
title: ""
author: "Giorgio Arcara"
date: ""
#format: revealjs
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
% TITLE SLIDE
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{7.2 - The pragmatic utility of thresholds}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{

\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## A quick recap

Normative data and thresholds helps us in interpreting scores

However, _per se_ \textcolor{red}{they do not tell anything about the existence of a pathological condition or a problem}


```{r}
#initialize a quato template for resuse
knitr::opts_template$set(
  myplot = list(
    echo = FALSE,
    fig.show = "asis",
    dpi = 100,
    fig.width = 7,
    fig.height = 5
  )
)

```

## What is the meaning of a score?

Suppose you have a score of **15 out of 30** on a test that aims at measuring anxiety, developed in the context of Classical Test Theory.

How to make a first interpretation of this score?

<!-- QUIZ HERE
- Suppose you have a score of 15 out of 30 in an Anxiety test.
- The person has an intermediate level of anxiety
- The person has an high level of anxiety
- The person has a low level of anxiety
- With the available information I cannot know the level of Anxiety.
-->

```{r}
#initialize a quato template for resuse
knitr::opts_template$set(
  myplot = list(
    echo = FALSE,
    fig.show = "asis",
    dpi = 100,
    fig.width = 7,
    fig.height = 5
  )
)
```

## Percentile and clinical relevance
\centering
Is there clinical relevance when an observed value falls at extreme percentiles or exceeds a threshold?


## Distribution of Anxiety score in healty participants
```{r}
#| opts.label: myplot

my_score=22
set.seed(42)
hist(rnorm(100, mean=16, sd=3), ylim=c(0,22), xlim=c(0,30), xlab="Score", ylab="N. of Observations", breaks=20, main="Anxiety Test")
segments(my_score, 0, my_score, 1, lwd=3, col="blue")
abline(v=my_score, lty=2, col="blue")

```

## Distribution of height in a normative sample
```{r}
#| opts.label: myplot

my_score=190
set.seed(100)
hist(rnorm(100, mean=175, sd=8), ylim=c(0,22), xlim=c(100,200), xlab="Score", ylab="N. of Observations", breaks=20, main="Height")
segments(my_score, 0, my_score, 1, lwd=3, col="blue")
abline(v=my_score, lty=2, col="blue")


```


## Percentiles and thresholds

_It helps us estimate the likelihood of observing a specific score (or equal or lower) in the normative data, from which we draw further inferences._

Thresholds are a further help to transform from a continuous distribution (percentiles) to a limited number of **arbitrary** categories.

```{=latex}
\begin{tikzpicture}[
  node distance=4cm,
  every node/.style={
    draw,
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center
  },
  arrow/.style={->, thick}
]

% Nodes (right to left)
\node (raw) {Raw score};

\node (perc) [right of=raw] {Percentile};

\node (cut) [right of=perc] {Comparison with\\thresholds/cut-off};

% Arrows
\draw [arrow] (raw) -- (perc);
\draw [arrow] (perc) -- (cut);

\end{tikzpicture}
```

## Paradox of the heap

```{=latex}
\begin{columns}[c]
\column{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/poor.png}
\end{figure}
\column{0.5\textwidth}
\pause
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/rich.png}
\end{figure}
\end{columns}
```

## Paradox of the heap

- **Premise 1**: If a person has 0 coin is "poor"

- **Premise 2**: If a person that is "poor" gain 1 coin is not rich
\pause

**Paradoxical conclusion**: if your apply the principle of Premise 2 a billion of time, the person should be still poor.

\pause
_Whenever a continuous variable is mapped onto vague categories, arbitrary thresholds create borderline cases that can give rise to Heap-type paradoxes._



## Thresholds based on statistical properties


If the threshold is based on statistical properties, it is a just convenient way to cluster data, but (often) based on arbitrary transformation of continuous percentile

## Thresholds and Validity

If we a support to the interpretation of thresholds, we should ascertain the _validity_ of this interpretation (see slides on **kwnon-groups validity**). 
\pause

However if we have **known groups** (e.g. people who are admitted to a psychiatric ward vs people who are not admitted at psychiatric ward), we can find optimal thresholds to **distinguish** between groups (see lectures of next weeks)



## Why we use arbitrary threholds?

If there are problems with thresholds, why don't we just use percentiles?

\pause
- There are cases in which we don't have clear "known" groups

\pause 
- They have **pragmatic utility**, and they can help us to make clinical choices (e.g. I will use this therapy for more severe cases)

## A hypothetical neuropsychological test (Test A)

| Task                     | Raw Score | Percentile | Cut-off |
|--------------------------|-----------|------------|---------|
| Memory                  | 45        | 60         |         |
| Attention               | 11        | 6         |       |
| Language                | 50        | 75         |         |
| Executive Functions     | 2        | 5         |        |
| Visuospatial            | 40        | 50         |         |
| Processing Speed        | 14        | 3         |        |

---

## A hypothetical neuropsychological test (Test B)

| Task                     | Raw Score | Percentile | Cut-off |
|--------------------------|-----------|------------|---------|
| Memory                  | 45        | 60         |         |
| Attention               | 11        | 6         |       |
| Language                | 50        | 75         |         |
| Executive Functions     | 2        | 5         | *       |
| Visuospatial            | 40        | 50         |         |
| Processing Speed        | 14        | 3         | *       |


## Take-home message

Thresholds based on normative data (i.e., derived only from healthy) are useful, but often they are not a proof to belonging to a specific clinical category.

This is because more often **we built** the categories arbitrarily, transforming a probability into groups.

Classification from norms should interpret them as a **potential warning sign** \includegraphics[scale=0.01]{Figures/triangle.png} that deserve further investigation.







