---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
    keep-md: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{9.1 - Normative data vs. Known-Groups}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## What we discussed so far

```{=latex}
\begin{columns}[c]
\column{0.5\textwidth}

\begin{figure}
\includegraphics[width=0.8\textwidth]{Figures/group_1.png}
\end{figure}
\onslide<2->{
\column{0.8\textwidth}
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/group_2.png}
}
\end{figure}
\end{columns}
\pause

In case of \textbf{kwnown-groups} other methods can be used (see also lectures on Validity)
```



## Two Ways of Deciding "Something deserves attention"

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Normative-based}\\[0.5em]
Compare the individual against a \emph{reference population}.\\[0.5em]
\small ``Your score is unusually low compared to people like you.''
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Known-groups}\\[0.5em]
Compare the individual against a \emph{known group}.\\[0.5em]
\small ``Your score resembles those of people \emph{with} the condition.''
\end{column}
\end{columns}

## Some terminology

In known-group testing we need to define two groups:

- **Condition of interest**:  the group we want to \emph{detect} (e.g., a clinical diagnosis).

- **Gold standard**:  the definitive, authoritative way of establishing who truly belongs to that group.

\vfill
\small
In known-group validity, the gold standard is the reference against which the test is validated.

## What Is a Gold Standard?

```{=latex}
\begin{tikzpicture}[node distance=1.5cm, auto,
  box/.style={rectangle, rounded corners, draw=black, fill=gray!15,
              text width=5cm, align=center, minimum height=1cm}]
\node[box] (gs) {\textbf{Gold Standard}\\(e.g., structured clinical interview, biopsy, long neuropsychological battery)};
\node[box, below=of gs] (test) {\textbf{Test under study}\\(shorter, cheaper, faster)};
\draw[-{Latex}, thick] (gs) -- node[right]{\small validates} (test);
\end{tikzpicture}
```

\vfill
\small
The gold standard is often expensive, time-consuming, or invasive <!-- — that is exactly why we need a shorter test. -->

## When Is a second group used in clinical psychology?

Two common situations in clinical/neuropsychological practice:

1. **Identifiable clinical group** — participants are recruited from a setting where the condition is established (e.g., patients admitted to a Psychiatric Ward, patients with a confirmed neurological diagnosis).

2. **Costly gold standard** — the definitive diagnosis requires a long procedure (e.g., full neuropsychological battery, multi-day observation), making a screening test valuable.

## Classification with known-group

Whenever a known-group is involved is useful to start thinking from **thresholds**



## Imagining the Screening Scenario

Suppose we administer a test to a **normative sample** only.

We can set a cut-off score below which we flag someone as potentially with clinical relevance.

\vfill

```{r}
#| echo: FALSE
#| fig.width: 6
#| fig.height: 3.5
#| fig.dpi: 100

set.seed(42)
x <- seq(-4, 4, length.out = 300)
y <- dnorm(x)

par(mar = c(4, 1, 3, 1))
plot(x, y, type = "l", lwd = 2, yaxt = "n", ylab = "",
     xlab = "Test Score", main = "Normative Distribution with Cut-off",
     col = "steelblue")

cutoff <- -1.65
polygon(c(x[x <= cutoff], cutoff), c(y[x <= cutoff], 0),
        col = rgb(1, 0.2, 0.2, 0.4), border = NA)
abline(v = cutoff, lty = 2, lwd = 2, col = "red")
text(cutoff - 0.1, 0.35, "cut-off\n(5th pct.)", col = "red", adj = 1, cex = 0.85)
text(-3, 0.05, "flagged", col = "darkred", cex = 0.8)
```

## True Negatives and False Positives

With a **normative reference only**, we can identify:
```{=latex}
\vfill

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{True Negative (TN)}\\[0.4em]
\small Score above cut-off \emph{and} person does not have the condition.\\
\textit{healthy person correctly flagged.}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{False Positive (FP)}\\[0.4em]
\small Score below cut-off \emph{but} person does not have the condition.\\
\textit{Healthy person incorrectly flagged.}
\end{column}
\end{columns}

\vfill
\small
%Without a group \emph{with} the condition, we \textbf{cannot} estimate True Positives or False %Negatives.
```

## Visualising TN and FP

```{r}
#| echo: FALSE
#| fig.width: 6
#| fig.height: 3.8
#| fig.dpi: 100

set.seed(42)
x   <- seq(-4, 4, length.out = 400)
y   <- dnorm(x)
cut <- -1.65

par(mar = c(4, 1, 3, 1))
plot(x, y, type = "l", lwd = 2, yaxt = "n", ylab = "",
     xlab = "Test Score", main = "Normative Sample: TN and FP",
     col = "steelblue")

# FP region (left of cut-off)
polygon(c(x[x <= cut], cut), c(y[x <= cut], 0),
        col = rgb(1, 0.3, 0.3, 0.5), border = NA)
# TN region (right of cut-off)
polygon(c(cut, x[x >= cut]), c(0, y[x >= cut]),
        col = rgb(0.2, 0.6, 1, 0.3), border = NA)

abline(v = cut, lty = 2, lwd = 2, col = "red")
legend("topright", legend = c("True Negative (TN)", "False Positive (FP)"),
       fill = c(rgb(0.2, 0.6, 1, 0.5), rgb(1, 0.3, 0.3, 0.5)), cex = 0.8)
```

## The Limitation of Normative Testing Alone

With only a normative sample we can estimate:

- **Specificity** (partially): how often do healthy people score above the cut-off?

\pause

But we **cannot** answer:

- How often does the test \emph{correctly identify} people with the condition?
- How many people with the condition will be \emph{missed}?

\vfill
\pause
\small
<!--
$\Rightarrow$ To answer these questions we need a group \textbf{with} the condition of interest.
-->

## From Norms to Known-groups: the Logic

```{=latex}
\begin{tikzpicture}[node distance=1.4cm,
  grp/.style={rectangle, rounded corners, draw, fill=gray!10,
              text width=3.8cm, align=center, minimum height=1cm}]
\node[grp] (norm) {Normative sample\\(no condition)};
\node[grp, right=2cm of norm] (clin) {Clinical sample\\(condition present)};
\node[grp, below=1.2cm of norm] (fp) {FP / TN estimable};
\node[grp, below=1.2cm of clin] (tp) {TP / FN estimable};
\draw[-{Latex}] (norm) -- (fp);
\draw[-{Latex}] (clin) -- (tp);
\draw[{Latex}-{Latex}, dashed] (norm) -- node[above]{\small compare} (clin);
\end{tikzpicture}
```

## Summary

| Concept | Definition |
|---|---|
| **Condition of interest** | The clinical state we aim to detect |
| **Gold standard** | Authoritative diagnostic reference |
| **True Negative** | Healthy person, correctly classified |
| **False Positive** | Healthy person, incorrectly classified |

\vfill
\small
Next: what happens when we \emph{also} have a group with the condition?
