---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{9.2 - Sensitivity, Specificity, and the ROC Curve}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## A short recap (normative data)

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.5
#| fig.dpi: 100

x <- seq(0, 16, length.out = 400)
y_ctrl <- dnorm(x, mean = 10, sd = 1)
y_clin <- dnorm(x, mean = 6, sd = 1)

par(mar = c(4, 1, 3, 1))
plot(x, y_ctrl, type = "l", lwd = 2, col = "steelblue",
     yaxt = "n", ylab = "", xlab = "Test Score",
     main = "Normative Distribution", ylim = c(0, 0.45), xlim=c(0,16))
#lines(x, y_clin, lwd = 2, col = "#FFAC1C")
abline(v = 8, lty = 2, lwd = 2, col = "gray40")
legend("topright",
       legend = c("Normative (controls)"),
       col = c("steelblue"), lwd = 2, cex = 0.85)
text(8.1, 0.42, "cut-off", col = "gray30", cex = 0.8, adj = 0)
```


## Adding the Clinical Group

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.5
#| fig.dpi: 100

x <- seq(0, 16, length.out = 400)
y_ctrl <- dnorm(x, mean = 10, sd = 1)
y_clin <- dnorm(x, mean = 6, sd = 1)

par(mar = c(4, 1, 3, 1))
plot(x, y_ctrl, type = "l", lwd = 2, col = "steelblue",
     yaxt = "n", ylab = "", xlab = "Test Score",
     main = "Normative and Clinical Distributions", ylim = c(0, 0.45), xlim=c(0,16))
lines(x, y_clin, lwd = 2, col = "#FFAC1C")
abline(v = 8, lty = 2, lwd = 2, col = "gray40")
legend("topright",
       legend = c("Normative (controls)", "Clinical (condition)"),
       col = c("steelblue", "#FFAC1C"), lwd = 2, cex = 0.85)
text(8.1, 0.42, "cut-off", col = "gray30", cex = 0.8, adj = 0)
```



## The Four Cells

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
 & \textbf{Condition present} & \textbf{Condition absent} \\
\hline
\textbf{Test positive} & True Positive (TP) & False Positive (FP) \\
\hline
\textbf{Test negative} & False Negative (FN) & True Negative (TN) \\
\hline
\end{tabular}
\end{center}
```

\vfill
\small
A test \emph{positive} means the score falls in the region associated with the condition (typically below a cut-off in deficit-oriented tests).

## True Positives and False Negatives

With a clinical group we can now also estimate:

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{True Positive (TP)}\\[0.4em]
\small Score falls in the "impaired" region \emph{and} person has the condition.\\
\textit{Correct detection.}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{False Negative (FN)}\\[0.4em]
\small Score falls in the "normal" region \emph{but} person has the condition.\\
\textit{Missed case.}
\end{column}
\end{columns}


## Classifications

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.8
#| fig.dpi: 100

x <- seq(0, 16, length.out = 400)
y_c   <-  dnorm(x, mean = 10, sd = 1)  # controls
y_p   <- dnorm(x, mean = 6, sd = 1)   # patients
cut   <- 8

par(mar = c(4, 1, 3, 1))
plot(x, y_c, type = "n", yaxt = "n", ylab = "",
     xlab = "Test Score", main = "Sensitivity / Specificity Trade-off",
     ylim = c(0, 0.45))

# FN (patients above cut-off)
xi <- x[x >= cut]; polygon(c(xi, rev(xi)), c(y_p[x >= cut], rep(0, sum(x >= cut))),
        col = rgb(1, 0.6, 0, 0.5), border = NA)
# TP (patients below cut-off)
xi <- x[x <= cut]; polygon(c(xi, rev(xi)), c(y_p[x <= cut], rep(0, sum(x <= cut))),
        col = "#FFAC1C", border = NA)
# FP (controls below cut-off)
xi <- x[x <= cut]; polygon(c(xi, rev(xi)), c(y_c[x <= cut], rep(0, sum(x <= cut))),
        col = rgb(1, 0.2, 0.9, 0.35), border = NA)
# TN (controls above cut-off)
xi <- x[x >= cut]; polygon(c(xi, rev(xi)), c(y_c[x >= cut], rep(0, sum(x >= cut))),
        col = rgb(0.2, 0.5, 1, 0.3), border = NA)

lines(x, y_c, lwd = 2, col = "steelblue")
lines(x, y_p, lwd = 2, col = "#FF7518")
abline(v = cut, lty = 2, lwd = 2)

legend("topright", cex = 0.7,
       legend = c("Controls", "Patients", "TP", "FN", "FP", "TN"),
       col    = c("steelblue","#FFAC1C", rgb(0.8,0.1,0.1,0.7),
                  rgb(1,0.6,0,0.7), rgb(1,0.2,0.9,0.5), rgb(0.2,0.5,1,0.5)),
       lwd    = c(2,2,NA,NA,NA,NA),
       pch    = c(NA,NA,15,15,15,15))
```



## Sensitivity

$$\text{Sensitivity} = \frac{TP}{TP + FN}$$

\vfill

\begin{itemize}
\item Proportion of \textbf{people with the condition} correctly identified by the test.
\item A \textbf{high sensitivity} means few cases are missed (low FN).
\item Also called the \emph{True Positive Rate}.
\end{itemize}

\vfill
\small
Critical when missing a case is costly (e.g., failing to detect a serious illness).

## Specificity

$$\text{Specificity} = \frac{TN}{TN + FP}$$

\vfill

\begin{itemize}
\item Proportion of \textbf{people without the condition} correctly classified.
\item A \textbf{high specificity} means few false alarms (low FP).
\item $1 - \text{Specificity}$ = False Positive Rate.
\end{itemize}

\vfill
\small
Critical when a false positive triggers costly or harmful follow-up procedures.


## Moving the Threshold Changes Everything

```{r}
#| echo: FALSE
#| fig.width: 7
#| fig.height: 4
#| fig.dpi: 100

x   <- seq(-2, 8, length.out = 500)
y_c <- dnorm(x, mean = 4, sd = 1)
y_p <- dnorm(x, mean = 2, sd = 1)

scenarios <- list(
  list(cut = 2.5, label = "High sensitivity\n(low threshold)"),
  list(cut = 3.0, label = "Balanced"),
  list(cut = 3.6, label = "High specificity\n(high threshold)")
)

par(mfrow = c(1, 3), mar = c(3, 1, 3.5, 1))
for (s in scenarios) {
  cut <- s$cut
  se_v <- round(pnorm(cut, 2, 1), 2)
  sp_v <- round(1 - pnorm(cut, 4, 1), 2)

  plot(x, y_c, type = "l", lwd = 2, col = "steelblue",
       yaxt = "n", ylab = "", xlab = "",
       main = s$label, ylim = c(0, 0.45), cex.main = 1.2)
  lines(x, y_p, lwd = 2, col = "tomato")
  abline(v = cut, lty = 2, lwd = 2, col = "gray30")

  # FN shading
  xi <- x[x > cut]
  polygon(c(xi, rev(xi)), c(y_p[x > cut], rep(0, sum(x > cut))),
          col = rgb(1, 0.6, 0, 0.45), border = NA)
  # FP shading
  xi <- x[x <= cut]
  polygon(c(xi, rev(xi)), c(y_c[x <= cut], rep(0, sum(x <= cut))),
          col = rgb(0.8, 0.2, 0.8, 0.35), border = NA)

  text(-1, 0.40, sprintf("Se=%.2f\nSp=%.2f", se_v, sp_v), cex = 1, adj = 0)
}
```

## The ROC Curve: Concept

As we move the cut-off across all possible values, we trace a curve:

- **x-axis**: False Positive Rate = $1 - \text{Specificity}$
- **y-axis**: True Positive Rate = Sensitivity

\vfill
This is the **Receiver Operating Characteristic (ROC) curve**.

\vfill
\small
Originally developed in signal detection theory (radar); now standard in medical/clinical testing.

## Plotting the ROC Curve

```{r}
#| echo: FALSE
#| fig.width: 4.5
#| fig.height: 4.5
#| fig.dpi: 100
#| out-width: "75%"
#| fig-align: "center"

cuts  <- seq(-2, 8, length.out = 300)
sens  <- pnorm(cuts, mean = 2, sd = 1)        # P(X < cut | patient)
fpr   <- 1 - pnorm(cuts, mean = 4, sd = 1)   # P(X < cut | control)

# Sort by FPR for clean plotting
ord  <- order(fpr)
#fpr  <- fpr[ord]; sens <- sens[ord]

par(mar = c(4, 4, 3, 1))
plot(1-fpr, sens, type = "l", lwd = 2, col = "tomato",
     xlab = "1 - Specificity (FPR)", ylab = "Sensitivity (TPR)",
     main = "ROC Curve", xlim = c(0,1), ylim = c(0,1))

abline(0, 1, lty = 2, col = "gray60")

# AUC via trapezoid rule
auc <- sum(diff(fpr[ord]) * (head(sens[ord],-1) + tail(sens[ord],-1)) / 2) # note that sens and spec are ordered.

legend("bottomright", legend = sprintf("AUC = %.2f", auc),
       bty = "n", cex = 0.9)
```

## Area Under the Curve (AUC)

The **AUC** summarises the overall discrimination ability of the test:

| AUC | Interpretation |
|-----|----------------|
| 0.5 | No discrimination (chance) |
| 0.70–0.79 | Acceptable |
| 0.80–0.89 | Good |
| $\geq$ 0.90 | Excellent |

\vfill
\small
AUC = probability that a randomly chosen patient scores higher (more impaired) than a randomly chosen control.

## Choosing the Optimal Threshold (Cut-off)

The ROC curve helps select a threshold balancing sensitivity and specificity:

- **Youden index**: maximise $\text{Sensitivity} + \text{Specificity} - 1$
- **Clinical priority**: weight sensitivity more if missing cases is costly; weight specificity more if false positives are costly.

\vfill
\small
There is no universally optimal cut-off: the choice depends on the clinical context.

## The Youden Index Is Not Always the Right Goal

The Youden Index maximises Se + Sp $- 1$, treating \textbf{false positives and false negatives as equally costly}.

\vfill

But in clinical practice they rarely are:

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{FN is costly}\\[0.3em]
\small Missing a serious condition (e.g., psychosis, early dementia) $\Rightarrow$ delayed treatment, harm to patient.\\[0.5em]
$\Rightarrow$ Prefer \textbf{high sensitivity}, accept more FP.
\end{column}
\begin{column}{0.48\textwidth}
\textbf{FP is costly}\\[0.3em]
\small Incorrect labelling (e.g., stigma, unnecessary follow-up, expensive procedures) $\Rightarrow$ harm to healthy person.\\[0.5em]
$\Rightarrow$ Prefer \textbf{high specificity}, accept more FN.
\end{column}
\end{columns}



<!--

## Weighted Criteria for Threshold Selection

If we can assign a \textbf{relative cost} to FP and FN, the optimal threshold is:

$$\text{maximise} \quad w_1 \cdot \text{Sensitivity} + w_2 \cdot \text{Specificity}$$

where $w_1$ and $w_2$ reflect the relative harm of each error type.

\vfill

\begin{itemize}
\item Youden is a special case: $w_1 = w_2 = 1$.
\item In screening for a serious condition: $w_1 \gg w_2$.
\item In confirmation (ruling in): $w_2 \gg w_1$.
\end{itemize}

\vfill
\small
The choice of weights is a \textbf{clinical and ethical} decision, not a statistical one.

## Generalisability of the Threshold

Even after a threshold is optimised and published, it may not generalise:

\begin{itemize}
\item The clinical and normative samples used for validation may not match the \textbf{target population}.
\item \textbf{Age, education, cultural background} can shift the score distributions.
\item The \textbf{severity mix} of the clinical sample (mild vs.\ severe cases) affects where the optimal threshold falls.
\end{itemize}

\vfill
\small
A threshold optimised on inpatients may perform poorly in a community outpatient setting.

-->




## Summary

| Index | Formula | What it tells us |
|---|---|---|
| **Sensitivity** | TP / (TP+FN) | Ability to detect the condition |
| **Specificity** | TN / (TN+FP) | Ability to exclude the condition |
| **AUC** | Area under ROC | Overall discrimination |

\vfill
\small