---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{9.3 - ROC, samples, and psychometrics}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## Sensitivity and Specificity Are Statistics Too

Sensitivity and specificity are estimated from samples: both the \textbf{normative} sample and the \textbf{clinical} sample.

\vfill

Like any statistic, they carry \textbf{sampling uncertainty}:

\begin{itemize}
\item A study with 20 patients may estimate Sensitivity $= 0.80$, but the true value could plausibly be anywhere from 0.55 to 0.95.
\item The \textbf{optimal threshold} on the ROC curve is found from the same data — it inherits the same uncertainty.
\end{itemize}

\vfill
\small
The sample–population distinction we applied to normative data applies \emph{equally} here.

## Thresholds of Discrimination Require Large Samples

The ROC curve, and the threshold derived from it, is estimated from finite samples.
With small samples, the curve is \textbf{unstable}: different samples from the same population yield different optimal thresholds.

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.6
#| fig.dpi: 100

set.seed(21)

roc_from_sample <- function(n) {
  pts  <- rnorm(n, mean = 2, sd = 1)
  ctrl <- rnorm(n, mean = 4, sd = 1)
  cuts <- seq(-1, 7, length.out = 200)
  tpr  <- sapply(cuts, function(c) mean(pts  <= c))
  fpr  <- sapply(cuts, function(c) mean(ctrl <= c))
  list(fpr = fpr, tpr = tpr)
}

par(mar = c(4, 4, 3, 1))
plot(NULL, xlim = c(0,1), ylim = c(0,1),
     xlab = "1 - Specificity (FPR)", ylab = "Sensitivity (TPR)",
     main = "ROC Curves from 20 Simulated Samples (n = 25 each)")
for (i in 1:20) {
  r <- roc_from_sample(25)
  ord <- order(r$fpr)
  lines(r$fpr[ord], r$tpr[ord], col = rgb(0.8, 0.2, 0.2, 0.25), lwd = 1)
}
# population ROC
cuts <- seq(-1, 7, length.out = 400)
tpr0 <- pnorm(cuts, 2, 1)
fpr0 <- pnorm(cuts, 4, 1)
ord  <- order(fpr0)
lines(fpr0[ord], tpr0[ord], col = "black", lwd = 2)
abline(0, 1, lty = 2, col = "gray60")
legend("bottomright", legend = c("Population ROC", "Sample ROCs (n=25)"),
       col = c("black", rgb(0.8,0.2,0.2,0.6)), lwd = c(2, 1), cex = 0.8)
```

<!--
## Confidence Intervals for Sensitivity and Specificity

At any given threshold, both indices are \textbf{proportions} — and proportions have confidence intervals.

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.8
#| fig.dpi: 100

ns      <- c(15, 25, 40, 60, 100, 200, 400)
true_se <- 0.80
hw      <- qnorm(0.975) * sqrt(true_se * (1 - true_se) / ns)

par(mar = c(4, 4, 3, 1))
plot(ns, hw, type = "b", pch = 19, col = "tomato", lwd = 2,
     xlab = "Sample size (clinical group)",
     ylab = "±95% CI half-width",
     main = "Precision of Sensitivity Estimate vs Sample Size\n(true Se = 0.80)",
     ylim = c(0, 0.35))
abline(h = 0.05, lty = 2, col = "gray50")
text(280, 0.068, "±0.05 precision threshold", col = "gray40", cex = 0.8)

# Annotate a few points
for (i in c(1, 3, 5, 7)) {
  text(ns[i], hw[i] + 0.018,
       sprintf("n=%d\n±%.2f", ns[i], hw[i]),
       cex = 0.65, col = "tomato")
}
```

-->


## Generalisability of the Threshold

The method shown so far use single samples (one for healthy one for the people with the condition of interest).

As such, it suffers the same limitation of methods using single groups (e.g. z-scores, non parametric percentiles, etc.)
\pause

Some extensions of ROC analysis include also demographic variables, via regression models.

## What about other Psychometric properties?

In general thresholds from ROC analysis (so for Known-Group Validity) does not incorporate directly other measurement properties, e.g. Reliability.
\pause

You should always prefer tests with **better psychometric properties**.

## Summary

\begin{itemize}
\item Sensitivity, specificity, and the derived threshold are \textbf{sample estimates}: they carry uncertainty.
\item Reliable discrimination thresholds require \textbf{large clinical and normative samples}.
\item The demographic variables of the samples used in the ROC analysis should be taken into account.
\item Good psychometric properties (e.g. Reliability), is desirable also with tests for discriminating between groups.
\end{itemize}

