---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{7.4 - PPV, NPV, and Bayes' Theorem}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## The Question Clinicians Actually Ask

Sensitivity and specificity answer questions about \emph{known groups}:

\begin{itemize}
\item $P(\text{test} + \mid \text{condition present})$ = Sensitivity
\item $P(\text{test} - \mid \text{condition absent})$ = Specificity
\end{itemize}

\pause
\vfill

But in practice the clinician asks the \textbf{inverse}:

\begin{center}
\large $P(\text{condition present} \mid \text{test}+) = \;?$
\end{center}

\vfill
\small
These are \textbf{not} the same probability — and confusing them is one of the most common errors in clinical reasoning.

## PPV and NPV: The Clinician's Indices

$$\text{PPV} = P(\text{condition} \mid \text{test}+) = \frac{TP}{TP + FP}$$

$$\text{NPV} = P(\text{no condition} \mid \text{test}-) = \frac{TN}{TN + FN}$$

\vfill

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{PPV}: if the test is positive, how likely is the condition?
\end{column}
\begin{column}{0.48\textwidth}
\textbf{NPV}: if the test is negative, how likely is the person healthy?
\end{column}
\end{columns}

\vfill
\small
Unlike sensitivity and specificity, PPV and NPV \textbf{depend on prevalence} — they change as the proportion of cases in the tested group changes.

## PPV and Bayes' Theorem Are the Same Thing

PPV is simply Bayes' theorem applied to a test result:

$$\underbrace{P(\text{condition} \mid \text{test}+)}_{\text{PPV}} = \frac{\overbrace{P(\text{test}+ \mid \text{condition})}^{\text{Sensitivity}} \times \overbrace{P(\text{condition})}^{\text{Prevalence}}}{P(\text{test}+)}$$

\vfill

where $P(\text{test}+) = \text{Se} \times \text{prev} + (1-\text{Sp}) \times (1-\text{prev})$.

\vfill
\small
\textbf{Prevalence} plays the role of the \emph{prior probability} — what we believed before the test. The test result then \emph{updates} this belief.

## A Concrete Example

Test: Sensitivity $= 0.90$, Specificity $= 0.95$. \quad Prevalence $= 1\%$.

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & \textbf{Condition +} & \textbf{Condition --} & \textbf{Total} \\
\hline
\textbf{Test +} & 9 \,(TP) & 50 \,(FP) & 59 \\
\hline
\textbf{Test --} & 1 \,(FN) & 940 \,(TN) & 941 \\
\hline
\textbf{Total} & 100 & 9\,900 & 10\,000 \\
\hline
\end{tabular}
\end{center}
```

\vfill

$$\text{PPV} = \frac{9}{59} \approx \mathbf{15\%} \qquad \text{NPV} = \frac{940}{941} \approx \mathbf{99.9\%}$$

\small
Despite excellent test properties, a positive result is \emph{wrong} 85\% of the time when prevalence is low.

## Prevalence as the Prior: A Visual

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 4.0
#| fig.dpi: 100

prev <- seq(0.01, 0.60, by = 0.005)
se   <- 0.90
sp   <- 0.95

ppv <- (se * prev) / (se * prev + (1 - sp) * (1 - prev))
npv <- (sp * (1 - prev)) / (sp * (1 - prev) + (1 - se) * prev)

par(mar = c(4, 4, 3, 1))
plot(prev, ppv, type = "l", lwd = 2, col = "tomato",
     xlab = "Prior probability (prevalence)",
     ylab = "Posterior probability",
     main = "Bayesian Update: PPV and NPV vs Prior\n(Se = 0.90, Sp = 0.95)",
     ylim = c(0, 1))
lines(prev, npv, lwd = 2, col = "steelblue")
abline(0, 1, lty = 3, col = "gray70")

# Annotate example points
points(0.01, ppv[which.min(abs(prev - 0.01))], pch = 19, col = "tomato", cex = 1.2)
points(0.30, ppv[which.min(abs(prev - 0.30))], pch = 19, col = "tomato", cex = 1.2)
text(0.02, 0.17, "prev=1%\nPPV≈15%", col = "tomato", cex = 0.72, adj = 0)
text(0.31, 0.83, "prev=30%\nPPV≈79%", col = "tomato", cex = 0.72, adj = 0)

legend("right",
       legend = c("PPV (posterior | test+)",
                  "NPV (posterior | test−)",
                  "No update (diagonal)"),
       col = c("tomato", "steelblue", "gray70"),
       lty = c(1, 1, 3), lwd = 2, cex = 0.78)
```

## Prior Probability in Clinical Practice

Prevalence figures assume the patient is a \textbf{random draw from the population} — which is rarely true.

\vfill

The relevant prior is the \textbf{pre-test probability} in the specific clinical context:

\begin{itemize}
\item A patient self-referred after reading about dementia online $\Rightarrow$ low prior.
\item A patient referred from a memory clinic with functional complaints $\Rightarrow$ much higher prior.
\item Published prevalence figures may not match either context.
\end{itemize}

\vfill
\small
In psychological and neuropsychological testing, reliable base-rate estimates are often unavailable. The prior must be estimated from \textbf{clinical judgment}.

## Belief Updating: From Prior to Posterior

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 4.0
#| fig.dpi: 100

se <- 0.90; sp <- 0.95
priors <- c(0.05, 0.15, 0.30, 0.50, 0.70)

ppv_pts <- (se * priors) / (se * priors + (1 - sp) * (1 - priors))
npv_pts <- (sp * (1 - priors)) / (sp * (1 - priors) + (1 - se) * priors)

x_pos <- seq_along(priors)
par(mar = c(4, 4, 3, 1))
plot(x_pos, ppv_pts, type = "b", pch = 19, col = "tomato", lwd = 2,
     ylim = c(0, 1), xaxt = "n",
     xlab = "Prior probability of condition",
     ylab = "Posterior probability",
     main = "How the Prior Determines the Posterior\n(Se = 0.90, Sp = 0.95)")
lines(x_pos, npv_pts, type = "b", pch = 17, col = "steelblue", lwd = 2)
lines(x_pos, priors,  type = "b", pch = 1,  col = "gray50",    lwd = 1, lty = 3)
axis(1, at = x_pos, labels = paste0(priors * 100, "%"))
legend("right",
       legend = c("PPV (after positive)", "NPV (after negative)", "Prior (no test)"),
       col = c("tomato", "steelblue", "gray50"),
       lty = c(1, 1, 3), pch = c(19, 17, 1), lwd = 2, cex = 0.82)
```

## The Odds Form: Easier to Chain

Bayes' theorem has a compact \emph{odds} form, useful when combining multiple pieces of evidence:

$$\underbrace{\frac{P(C \mid +)}{P(\overline{C} \mid +)}}_{\text{posterior odds}} = \underbrace{\frac{\text{Se}}{1-\text{Sp}}}_{\text{LR}^+} \times \underbrace{\frac{P(C)}{P(\overline{C})}}_{\text{prior odds}}$$

\vfill

The ratio $\text{LR}^+ = \frac{\text{Se}}{1-\text{Sp}}$ is the \textbf{Likelihood Ratio} — it tells us how much a positive result multiplies the prior odds, \emph{independently of prevalence}.

\vfill
\small
\textbf{LR} $> 10$: strong evidence for the condition. \quad \textbf{LR} $< 0.1$: strong evidence against.

## Sequential Updating

Because Bayesian updating is multiplicative in odds, we can apply it \textbf{step by step} as new evidence accumulates:

```{=latex}
\begin{tikzpicture}[node distance=0.5cm and 1.0cm,
  ev/.style={rectangle, rounded corners, draw=gray!70, fill=gray!10,
             text width=2.6cm, align=center, minimum height=1.0cm, font=\small},
  arr/.style={-{Latex}, thick}]
\node[ev] (p0) {Clinical interview\\Prior: \textbf{20\%}};
\node[ev, right=1.0cm of p0] (p1) {Screening +\\Updated: \textbf{62\%}};
\node[ev, right=1.0cm of p1] (p2) {Full battery +\\Updated: \textbf{89\%}};
\node[ev, right=1.0cm of p2] (p3) {Family report\\Updated: \textbf{93\%}};
\draw[arr] (p0) -- (p1);
\draw[arr] (p1) -- (p2);
\draw[arr] (p2) -- (p3);
\node[below=0.4cm of p0, font=\tiny, text=gray] {LR context};
\node[below=0.4cm of p1, font=\tiny, text=gray] {LR$^+$= 6.0};
\node[below=0.4cm of p2, font=\tiny, text=gray] {LR$^+$= 4.5};
\node[below=0.4cm of p3, font=\tiny, text=gray] {LR$^+$= 2.0};
\end{tikzpicture}
```

\vfill
\small
No single source is decisive. Each piece of evidence — test score, interview, anamnesis, collateral information — is one update. The clinician's job is to \textbf{integrate} them.

## Summary

\scriptsize
\begin{tabular}{|p{2.4cm}|p{4.8cm}|}
\hline
\textbf{Concept} & \textbf{Key point} \\
\hline
PPV / NPV & Posterior probabilities; answer the clinician's question \\
\hline
Prevalence & Acts as the prior; context-specific, often unknown \\
\hline
Bayes' theorem & Formally connects prior, likelihood (Se/Sp), and posterior \\
\hline
Likelihood Ratio & Prevalence-free; quantifies the weight of a single result \\
\hline
Sequential updating & Each new source of evidence is one Bayesian update \\
\hline
\end{tabular}

\vfill
\small
Testing is not a binary verdict — it is a \textbf{progressive accumulation of evidence} that updates the probability of the condition.

