---
title: ""
author: "Giorgio Arcara"
date: ""
#format: revealjs
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
  \usepackage{booktabs}
  \usepackage{colortbl}
  \usepackage{xcolor}
---

<!-- the section below is shown only if rendering is beamer from latex. (and format:revealjs is commmented) -->

```{=latex}
% TITLE SLIDES
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{11.1 - Detecting Change over time}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{

\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## Many reasons to detect change over time

- Compare scores before and after a treatment \pause

- In a chronic disease monitoring scores over time to see if symptoms are improving/worsening \pause

- In a degenerative disease monitoring if performance is worsening or is stable \pause

- \ldots


## Is this change meaningful?

A patient completes an anxiety questionnaire at two timepoints:

```{=latex}
\vspace{0.8em}
\begin{center}
\begin{tabular}{lcc}
\toprule
 & \textbf{Time 1} & \textbf{Time 2} \\
\midrule
Anxiety score & 38 & 34 \\
\bottomrule
\end{tabular}
\end{center}

\pause
\vspace{0.8em}
The score dropped by \textbf{4 points}. Is the patient \emph{really} better?

%\vspace{0.5em}
%\textbf{Not necessarily.} All psychological tests have %measurement error. A difference of 9 points might simply reflect %noise --- not true change.
```

## The common (but wrong) approach

A frequent practice is to compare scores against **normative thresholds**, e.g., classifying a patient as "improved" because their score moved from the *clinical* to the *non-clinical* range.

\pause

```{=latex}
\vspace{0.5em}
\begin{alertblock}{Why this is wrong}
\vspace{0.2em}
Normative thresholds tell us where a score stands \emph{relative to a sample/population}, but not if the change between scores deserve attention. 
\end{alertblock}
```

## Change over time and reliability

In the context of Classical Test Theory, detecting change over time is tightly related to Test-retest reliability.

\pause
To detect a _meaningful_ change they test whether the \emph{change between two scores} exceeds what we would expect by chance due to measurement error.

## Two observations with a High-reliability test

```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "100%"


library(ONDAutils)

set.seed(10)
X = simulate_obs(Ts = 38, n = 30, rel=0.9, mu_true = 35, sd_true = 2, label = "High Reliability")
abline(h=38, col="blue", lwd=3)
legend("topright", legend=c("observed score", "true score"), pch=c(1, NA), lty=c(NA, 1), lwd=c(NA, 3), col=c("black", "blue"))

```

## Two observations with a Low-reliability test

```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "100%"


set.seed(10)
X = simulate_obs(Ts = 38, n = 30, rel=0.5, mu_true = 35, sd_true = 2, label = "Low Reliability")
abline(h=38, col="blue", lwd=3)
legend("topright", legend=c("observed score", "true score"), pch=c(1, NA), lty=c(NA, 1), lwd=c(NA, 3), col=c("black", "blue"))

```

## Is this a "real" change?

```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "100%"


set.seed(10)
X = simulate_obs(Ts = 38, n = 30, rel=0.5, mu_true = 35, sd_true = 2, label = "Low Reliability")
abline(h=38, col="blue", lwd=3)
legend("topright", legend=c("observed score", "true score"), pch=c(1, NA), lty=c(NA, 1), lwd=c(NA, 3), col=c("black", "blue"))

points(10, X[10],  pch=21, bg="orange",col="black", cex=2)

```

## Is this a "real" change?

```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "100%"


set.seed(10)
X = simulate_obs(Ts = 38, n = 30, rel=0.5, mu_true = 35, sd_true = 2, label = "Low Reliability")
abline(h=38, col="blue", lwd=3)
legend("topright", legend=c("observed score", "true score"), pch=c(1, NA), lty=c(NA, 1), lwd=c(NA, 3), col=c("black", "blue"))

points(10, X[10],  pch=21, bg="orange",col="black", cex=2)
points(11, X[11]-1,  pch=21, bg="orange",col="black", cex=2)
```
\small
\emph{A difference of two points $X_{t0}$ = 37, $X_{t1}$=39, seems negligible in this case.}
<!--

## Reliability and measurement error (CTT)

Within **Classical Test Theory (CTT)**, every observed score $X$ contains a true score $T$ and an error $E$:

$$X = T + E$$

The **Standard Error of Measurement** (SEM) quantifies the typical size of $E$:

$$\text{SEM} = SD \sqrt{1 - r_{xx}}$$

where $r_{xx}$ is the reliability coefficient of the test.
\pause

A \textbf{higher reliability} $\rightarrow$ smaller SEM $\rightarrow$ more confidence that observed change is real.
-->

## How to Build Normative Data for Change Detection
A group is tested twice (separated by a pre-specified time interval), during which **no true change is assumed to have occurred**.

Any fluctuations observed in the scores are therefore attributable to measurement error.

## How to solve?

One simile possibility is to calculate z-scores from the difference 
```{=latex}
\centering
$X_{t1} - X_{t0}$
```

## Normative data (raw score)
```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "90%"

par(mfrow=c(1,2))
set.seed(10)
hist(rnorm(100, mean=20), col="steelblue", bg="steelblue", main="Normative data (raw score)\n at t0", xlab=expression("X"))

set.seed(11)
hist(rnorm(100, mean=20), col="steelblue", bg="steelblue", main="Normative data (raw score)\n at t1", xlab=expression("X"))


```
\emph{Note the x-axis}


## Normative data (Difference score)

```{r} 
#| echo: FALSE
#| output: TRUE
#| fig.dpi: 100
#| fig.width: 6
#| fig.height: 4
#| out-width: "90%"


set.seed(10)
xt0 = rnorm(100, mean=20)
set.seed(11)
xt1 = rnorm(100, mean=20)

xdiff = xt1-xt0

hist(rnorm(100), col="steelblue", bg="steelblue", main="Normative data (difference score)\n t1 minus t0", xlab=expression("X"[t1] - "X"[t0]))
```
\emph{Note the x-axis. Using differences assumes "interval scale", see Lectures of Week 1}

## Assumption: interval scale
Imagine a test in which scores range from 0 to 30

\pause
\vspace{1em}
$X_{t0}$ = 25, $X_{t1}$ = 30. \pause  $\Delta = X_{t1}- X_{t0}$ = +5

\pause
\vspace{1em}
$X_{t0}$ = 15, $X_{t1}$ = 20 $\Delta = X_{t1}- X_{t0}$ = +5


## Summary

Percentiles derived from normative data answer the question:

*How likely is to observe this score within the reference sample/population?*

\pause
Our question, instead, is:

*How likely is it to observe this score at $t_1$, given the score observed at $t_0$?*

## Methods to detect change
\centering

See the next video lecture!


