---
title: ""
author: "Giorgio Arcara"
date: ""
#format: revealjs
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

```{=latex}
% TITLE SLIDE
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{7.5 - From Percentiles to Probabilities}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{

\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

```{r}
knitr::opts_template$set(
  myplot = list(
    echo = FALSE,
    fig.show = "asis",
    dpi = 100,
    fig.width = 7,
    fig.height = 5
  )
)
```

## Recap: what does a z-score do?

The **z-score** transforms a raw score into a standardized unit, expressing how far a patient's score is from the **mean of the normative sample**:

$$Z = \frac{(X - M)}{SD}$$

When the normative distribution is normal, this z-score maps directly to a **percentile**.

\vspace{1em}
But... what exactly is that percentile telling us?


## Percentiles: a description of a sample

A percentile rank tells us:

> *"What proportion of people **in the normative sample** scored below this value?"*

\vspace{1em}

For example: a score at the **5th percentile** means that 95% of the normative sample scored higher.

\vspace{1em}

\textbf{Key word: sample.}  
The percentile is a \emph{descriptive} summary of the norm group — not a statement about the population.


## Sample vs. Population

```{=latex}
\begin{center}
\begin{tikzpicture}[node distance=1.5cm]
  \node[draw, rounded corners, fill=gray!15, minimum width=4.5cm, minimum height=1.2cm] (pop) {\textbf{Population}: all healthy adults};
  \node[draw, rounded corners, fill=blue!15, minimum width=4.5cm, minimum height=1.2cm, below=of pop] (samp) {\textbf{Normative sample}: $n = 150$ adults};
  \node[draw, rounded corners, fill=orange!20, minimum width=4.5cm, minimum height=1.2cm, below=of samp] (pat) {\textbf{Patient}: 1 individual};
  \draw[->, thick] (pop) -- (samp) node[midway, right] {\small drawn from};
  \draw[->, thick, dashed] (samp) -- (pat) node[midway, right] {\small compared to};
  \draw[->, thick, dotted, bend right=60] (pop.west) to node[left] {\small \emph{what we want}} (pat.west);
\end{tikzpicture}
\end{center}
```

The percentile compares the patient to the **sample**.  
What we actually want is to compare the patient to the **population**.


## The problem: uncertainty around the percentile
\small
A normative sample is an **estimate** of the population — and estimates carry uncertainty.

\vspace{1em}

If the normative sample has $n = 50$, the confidence interval around a percentile rank is **very wide**. A score at the 5th percentile could plausibly be at the 2nd or the 12th when we account for sampling error.

\vspace{0.5em}

```{r}
#| opts.label: myplot
#| fig.height: 2

set.seed(42)
opt(par())
ns <- c(20, 50, 100, 200, 500)
# CI width for a proportion p=0.05
p <- 0.05
ci_width <- sapply(ns, function(n) 2 * 1.96 * sqrt(p*(1-p)/n) * 100)
barplot(ci_width, names.arg=ns, xlab="Normative sample size (n)",
        ylab="95% CI width (percentile points)",
        main="Uncertainty around the 5th percentile",
        col="steelblue", ylim=c(0, 20))
```


## Percentage vs. Probability

Two concepts that are easy to conflate but are fundamentally different:

\vspace{1em}

| | **Percentile rank** | **p-value** |
|---|---|---|
| What it is | Proportion of sample below a score | Probability under a population model |
| Logic | Descriptive | Inferential |
| Depends on | The sample | A statistical model |
| Answers | "Where in the sample?" | "How probable under $H_0$?" |

\vspace{1em}

A score at the **5th percentile** is not the same as $p = .05$.


## Why this matters clinically

Consider a patient who scores at the **5th percentile** on a memory test.

\vspace{1em}

Using a **percentile**: we say "95% of the norm group scored higher" — but this is silent about:

- How uncertain is the norm group estimate?
- How probable is this score given the *population*?
- Is this score truly *abnormal*, or just low-average given sampling noise?

\vspace{1em}

\textbf{The clinical interpretation requires an inferential, not merely descriptive, tool.}


## A single case: the classical approach

The classical approach treats the z-score as if the normative sample **were** the population:

$$Z = \frac{X - M}{SD} \sim \mathcal{N}(0,1)$$

This assumes $M$ and $SD$ are known **population parameters** — but they are only **sample estimates**.

\vspace{1em}

For large normative samples ($n > 500$) this approximation is reasonable.  
For typical neuropsychological norms ($n = 50$–$200$)... it is not.


## Crawford & Howell (1998): the modified $t$-test

Crawford & Howell proposed treating the normative sample mean and SD as **estimates**, not fixed values. The patient's score is tested using a **modified one-sample $t$-test**:

$$t = \frac{X - M}{SD \sqrt{\frac{n+1}{n}}}$$

with $df = n - 1$.

\vspace{1em}

This is a small but crucial correction: it explicitly acknowledges that $M$ and $SD$ are uncertain, inflating the standard error accordingly.


## What the Crawford $t$-test actually asks

The z-score asks:

> *"Where does this score fall in the normative sample?"*

\vspace{1em}

The Crawford $t$-test asks:

> *"Given that the normative sample is our best estimate of the population, what is the **probability** of observing a score this extreme in the population?"*

\vspace{1em}

This is an **inferential** question — and the one we actually care about clinically.


## Comparison: z-score vs. Crawford $t$-test

```{r}
#| opts.label: myplot
#| fig.height: 4

# Simulate: patient score X=8, normative sample M=12, SD=3
X <- 8; M <- 12; SD <- 3
ns <- seq(10, 300, by=5)

p_z <- pnorm((X - M) / SD)  # fixed, doesn't change with n

p_t <- sapply(ns, function(n) {
  t_val <- (X - M) / (SD * sqrt((n + 1) / n))
  pt(t_val, df = n - 1)
})

plot(ns, p_t * 100, type="l", lwd=2, col="steelblue",
     xlab="Normative sample size (n)",
     ylab="One-tailed p (as %)",
     main="p-value: z-score vs Crawford t-test",
     ylim=c(0, 15))
abline(h = p_z * 100, col="firebrick", lwd=2, lty=2)
legend("topright", legend=c("Crawford t-test", "z-score (fixed)"),
       col=c("steelblue","firebrick"), lwd=2, lty=c(1,2))
```

\small For small $n$, the Crawford $t$-test yields a less extreme (more honest) $p$-value.


## Beyond the point estimate: abnormality and rarity

Crawford et al. also distinguish two inferential questions:

\vspace{1em}

1. **Is the score statistically abnormal?**  
   → How improbable is this score given the population? (the modified $t$-test)

\vspace{0.5em}

2. **How rare is a deficit this large?**  
   → What proportion of the population scores at or below this level? (estimated with uncertainty)

\vspace{1em}

Both are probabilistic statements about the **population**, not descriptions of the sample.  
Both come with **confidence intervals** — the percentile alone does not.


## Putting it all together

```{=latex}
\begin{center}
\begin{tikzpicture}[node distance=1.2cm, every node/.style={font=\small}]
  \node[draw, fill=gray!10, rounded corners, minimum width=5cm] (q) {Clinical question: is this score abnormal?};
  \node[draw, fill=blue!10, rounded corners, minimum width=5cm, below=of q, align=center] (perc) {Percentile rank (z-score):\\descriptive, sample-bound};
  \node[draw, fill=green!10, rounded corners, minimum width=5cm, below=of perc, align=center] (craw) {Crawford $t$-test:\\ inferential, population-referenced};
  \node[draw, fill=orange!10, rounded corners, minimum width=5cm, below=of craw, align=center] (ci) {+ Confidence interval on rarity:\\honest uncertainty quantification};
  \draw[->, thick] (q) -- (perc) node[midway, right] {classical};
  \draw[->, thick] (q.west) to [bend right=40] node[left] {better} (craw.west);
  \draw[->, thick] (craw) -- (ci);
\end{tikzpicture}
\end{center}
```


## Practical recommendations

When interpreting a single-case neuropsychological score:

\vspace{0.5em}

- **Always report the normative sample size** — it determines how much to trust the percentile
- **Prefer inferential methods** (Crawford $t$-test) over raw percentile ranks when $n < 200$
- **Distinguish** "rare in this sample" (percentile) from "improbable in the population" ($p$-value)
- **Report confidence intervals** around estimates of abnormality
- **Be explicit** about whether you are making a descriptive or inferential claim

\vspace{0.5em}

\small \textit{Crawford, J.R., \& Howell, D.C. (1998). Comparing an individual's test score against norms derived from small samples. \textit{The Clinical Neuropsychologist}, 12(4), 482--486.}


## Summary

| Concept | Percentile / z-score | Crawford $t$-test |
|---|---|---|
| Logic | Descriptive | Inferential |
| Reference | Normative sample | Population (estimated) |
| Uncertainty | Ignored | Explicit |
| Output | Rank / proportion | Probability ($p$-value) |
| Best for | Large norm samples | Small/medium norm samples |

\vspace{1em}

\centering \textbf{The patient is not in your normative sample.}\\
\textit{Treat the norm group as an estimate — not as ground truth.}

-->