---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

::: {.content-visible when-format="beamer"}
```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{10.1 - PPV, NPV, and Likelihood Ratios}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## An example

Suppose you have a test with excellent Sensitity and Specitificy (e.g. 0.99 Sensitivity and 0.98 Specificity) to discriminate from people with OCD from people with OCD.
\pause

Suppose that the patient score below the thresholds and has a _positive_ result.
\pause

What would you conclude?

<!-- QUIZ -->

## The Question Clinicians Actually Ask

Sensitivity and specificity answer questions about \emph{known groups}:

\begin{itemize}
\item ``Given the patient \textbf{has} the condition, how likely is a positive test?''
\item ``Given the patient \textbf{does not have} the condition, how likely is a negative test?''
\end{itemize}

\pause
\vfill

But in practice the clinician asks the \textbf{opposite}:

\begin{center}
\large\textit{``Given a positive test, how likely is the condition?''}
\end{center}

\vfill
\small
This is what \textbf{Positive Predictive Value} answers.

## Positive and Negative Predictive Value

$$\text{PPV} = \frac{TP}{TP + FP} \qquad \text{NPV} = \frac{TN}{TN + FN}$$

\vfill

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{PPV}\\[0.4em]
\small Probability that the condition is \emph{present} given a positive test result.
\end{column}
\begin{column}{0.48\textwidth}
\textbf{NPV}\\[0.4em]
\small Probability that the condition is \emph{absent} given a negative test result.
\end{column}
\end{columns}

\vfill
\small
Unlike sensitivity and specificity, PPV and NPV \textbf{depend on prevalence}.

## Prevalence Drives PPV

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 4
#| fig.dpi: 100

prev <- seq(0.01, 0.50, by = 0.005)
se   <- 0.85
sp   <- 0.90

ppv  <- (se * prev) / (se * prev + (1 - sp) * (1 - prev))
npv  <- (sp * (1 - prev)) / (sp * (1 - prev) + (1 - se) * prev)

par(mar = c(4, 4, 3, 1))
plot(prev, ppv, type = "l", lwd = 2, col = "tomato",
     xlab = "Prevalence", ylab = "Predictive Value",
     main = "PPV and NPV as a Function of Prevalence\n(Sensitivity = 0.85, Specificity = 0.90)",
     ylim = c(0, 1))
lines(prev, npv, lwd = 2, col = "steelblue")
abline(v = 0.05, lty = 2, col = "gray50")
text(0.06, 0.15, "5%\nprevalence", col = "gray40", cex = 0.75, adj = 0)
legend("right", legend = c("PPV", "NPV"),
       col = c("tomato", "steelblue"), lwd = 2, cex = 0.85)
```

## A Striking Example

Suppose a test has Sensitivity = 0.90 and Specificity = 0.95.

We screen a population where the condition affects **1 in 100** people.

\vfill

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & \textbf{Condition +} & \textbf{Condition --} & \textbf{Total} \\
\hline
\textbf{Test +} & 9 (TP) & 50 (FP) & 59 \\
\hline
\textbf{Test --} & 1 (FN) & 940 (TN) & 941 \\
\hline
\textbf{Total} & 100 & 9900 & 10\,000 \\
\hline
\end{tabular}
\end{center}
```

\vfill

$$\text{PPV} = \frac{9}{59} \approx \mathbf{15\%}$$

\small
A positive result is \emph{wrong} 85\% of the time — despite excellent test properties!

## The Limitation of Prevalence Figures

Prevalence estimates assume the person is a \textbf{random draw from the population}.

\vfill

In a clinical context this rarely holds:

\begin{itemize}
\item A patient referred to a memory clinic has a \emph{higher} pre-test probability of MCI than the general population.
\item A patient screened opportunistically at a GP visit has a \emph{lower} pre-test probability.
\end{itemize}

\vfill
\small
PPV and NPV computed from published prevalence figures may be \textbf{misleading} if the clinical context differs substantially.

## Prevalence in Psychological Testing

\begin{itemize}
\item Epidemiological prevalence estimates are available for some psychiatric disorders (e.g., major depression, schizophrenia), but often with wide confidence intervals.
\item For many neuropsychological constructs (e.g., ``executive dysfunction'', ``memory deficit'') \textbf{no reliable prevalence figure exists}.
\item Base rates shift dramatically across settings: community, primary care, specialist clinic, inpatient ward.
\end{itemize}

\vfill
\small
In practice, the clinician's \textbf{clinical judgment} — not a published prevalence — often serves as the prior probability. This motivates the Bayesian approach (see next group).

## Likelihood Ratios: A Prevalence-Free Alternative

Likelihood Ratios (LR) express how much a test result \emph{shifts} the probability of the condition — without requiring a prevalence estimate.

$$LR^{+} = \frac{\text{Sensitivity}}{1 - \text{Specificity}} \qquad LR^{-} = \frac{1 - \text{Sensitivity}}{\text{Specificity}}$$

\vfill

\begin{itemize}
\item $LR^{+}$: how much more likely a positive result is in someone \emph{with} vs.\ \emph{without} the condition.
\item $LR^{-}$: how much more likely a negative result is in someone \emph{with} vs.\ \emph{without} the condition.
\end{itemize}

## Interpreting Likelihood Ratios

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.5
#| fig.dpi: 100

lrp <- c(1, 2, 5, 10, 20)
labs <- c("No change\n(LR=1)", "Slight\n(LR=2)", "Moderate\n(LR=5)",
          "Strong\n(LR=10)", "Very strong\n(LR=20)")
cols <- c("gray70", "#f0c060", "#e08030", "#c04020", "#801010")

par(mar = c(3, 5, 3, 1))
barplot(lrp, names.arg = labs, col = cols, border = NA,
        horiz = FALSE, ylab = "LR+",
        main = "Interpreting LR+: Rule of Thumb",
        cex.names = 0.75)
abline(h = 1, lty = 2, col = "gray40")
```

\small
Rule of thumb: LR $>$ 10 provides strong evidence; LR $<$ 0.1 provides strong evidence \emph{against} the condition.

## LR: Stable Across Prevalence

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.8
#| fig.dpi: 100

se <- 0.85; sp <- 0.90
lrp_val <- se / (1 - sp)

prev  <- seq(0.01, 0.60, by = 0.01)
ppv_v <- (se * prev) / (se * prev + (1 - sp) * (1 - prev))

# Posterior odds = LR * prior odds
prior_odds <- prev / (1 - prev)
post_odds  <- lrp_val * prior_odds
post_prob  <- post_odds / (1 + post_odds)

par(mar = c(4, 4, 3, 1))
plot(prev, ppv_v, type = "l", lwd = 2, col = "tomato",
     xlab = "Prior probability (prevalence)",
     ylab = "Posterior probability",
     main = sprintf("LR+ = %.1f: Converting Prior to Posterior", lrp_val),
     ylim = c(0, 1))
lines(prev, post_prob, lwd = 2, lty = 2, col = "tomato")
abline(0, 1, lty = 3, col = "gray60")
legend("topleft", legend = c("PPV (exact)", "LR approximation", "No update (diagonal)"),
       col = c("tomato", "tomato", "gray60"),
       lty = c(1, 2, 3), lwd = 2, cex = 0.75)
```

## Summary: PPV, NPV, LR

\small
\begin{tabular}{|p{2.3cm}|p{3cm}|p{3cm}|}
\hline
\textbf{Index} & \textbf{What it tells us} & \textbf{Depends on prevalence?} \\
\hline
PPV & P(condition $|$ test +) & Yes \\
\hline
NPV & P(no condition $|$ test --) & Yes \\
\hline
LR+ & Evidence shift from positive result & No \\
\hline
LR-- & Evidence shift from negative result & No \\
\hline
\end{tabular}

\vfill
\small
Use LRs when prevalence is unknown or context-specific; use PPV/NPV when a reliable prevalence estimate is available and the clinical context matches.

:::
