---
title: ""
author: "Giorgio Arcara"
date: ""
format:
  beamer:
    theme: metropolis
    keep-tex: false
incremental: false
header-includes:
  \setbeamerfont{title}{size=\large}
  \usepackage{hyperref}
  \setbeamertemplate{footline}[frame number]
  \usepackage{tikz}
  \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
  \usetikzlibrary{overlay-beamer-styles}
---

::: {.content-visible when-format="beamer"}
```{=latex}
\title{{\fontsize{15pt}{16pt}\selectfont Applied Psychometrics}:\\ studying subclinical psychopathologies\\ \vspace{1.5em} \raggedright 
\emph{10.2 - Bayesian Reasoning in Clinical Testing}}
\author{Giorgio Arcara,\\ University of Padua \\ IRCCS San Camillo, Venice}

\titlegraphic{
\vspace*{7cm}
\includegraphics[scale=0.15]{Figures/LogoSanCamilloIRCCS_Unipd_alpha.png}
\begin{center}
\vspace{-2.2em}
\includegraphics[scale=0.15]{Figures/CC_license_3_0.png}
\end{center}
}
\date{2026}
\maketitle
```

## A Common Confusion

When a patient scores below the cut-off, it is tempting to say:

\begin{center}
\large ``The probability that this person has the condition is 85\%''\\
\small \textit{(if sensitivity = 0.85)}
\end{center}

\pause
\vfill

This is \textbf{wrong}.

\vfill

Sensitivity tells us:
$$P(\text{score below cut-off} \mid \text{condition present}) = 0.85$$

What we \emph{want} is:
$$P(\text{condition present} \mid \text{score below cut-off}) = \;?$$

## These Are Not the Same Thing

$$P(A \mid B) \neq P(B \mid A)$$

\vfill
```{=latex}
A vivid analogy:\\
\vspace{0.5cm}
$P(\text{fever} \mid \text{meningitis}) \approx 1$\\
\vspace{0.4cm}
$P(\text{meningitis} \mid \text{fever}) \ll 1$
{\scriptsize (fever is more often associated to minor pathologies)}
```

## These Are Not the Same Thing

$$P(A \mid B) \neq P(B \mid A)$$

```{=latex}
In testing:\\
\vspace{0.5cm}
$P(\text{low score} \mid \text{dementia}) \approx $ Sensitivity\\
\vspace{0.4cm}
$P(\text{dementia} \mid \text{low score}) = $ what the clinician wants \\
\pause $\to$ requires \textbf{Bayes' theorem}.
```

## Bayes' Theorem (Simply Stated)

$$P(\text{condition} \mid \text{result}) = \frac{P(\text{result} \mid \text{condition}) \times P(\text{condition})}{P(\text{result})}$$

\vfill

In Bayes theorem language:

\begin{center}
\textbf{Posterior} $\propto$ \textbf{Likelihood} $\times$ \textbf{Prior}
\end{center}

```{=latex}
\vfill
\small
The \emph{prior} is what we believed before the test.\\
The \emph{likelihood} is provided by sensitivity and specificity.\\
The \emph{posterior} is our updated belief after seeing the result.
```

## Likelihood Ratios: A Prevalence-Free Alternative

Likelihood Ratios (LR) express how much a test result \emph{shifts} the probability of the condition — without requiring a prevalence estimate.

$$LR^{+} = \frac{\text{Sensitivity}}{1 - \text{Specificity}} \qquad LR^{-} = \frac{1 - \text{Sensitivity}}{\text{Specificity}}$$

\vfill

\begin{itemize}
\item $LR^{+}$: how much more likely a positive result is in someone \emph{with} vs.\ \emph{without} the condition.
\item $LR^{-}$: how much more likely a negative result is in someone \emph{with} vs.\ \emph{without} the condition.
\end{itemize}

## Interpreting Likelihood Ratios

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.5
#| fig.dpi: 100

lrp <- c(1, 2, 5, 10, 20)
labs <- c("No change\n(LR=1)", "Slight\n(LR=2)", "Moderate\n(LR=5)",
          "Strong\n(LR=10)", "Very strong\n(LR=20)")
cols <- c("gray70", "#f0c060", "#e08030", "#c04020", "#801010")

par(mar = c(3, 5, 3, 1))
barplot(lrp, names.arg = labs, col = cols, border = NA,
        horiz = FALSE, ylab = "LR+",
        main = "Interpreting LR+: Rule of Thumb",
        cex.names = 0.75)
abline(h = 1, lty = 2, col = "gray40")
```

\small
Rule of thumb: LR $>$ 10 provides strong evidence; LR $<$ 0.1 provides strong evidence \emph{against} the condition.

<!--
## LR: Stable Across Prevalence

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 3.8
#| fig.dpi: 100

se <- 0.85; sp <- 0.90
lrp_val <- se / (1 - sp)

prev  <- seq(0.01, 0.60, by = 0.01)
ppv_v <- (se * prev) / (se * prev + (1 - sp) * (1 - prev))

# Posterior odds = LR * prior odds
prior_odds <- prev / (1 - prev)
post_odds  <- lrp_val * prior_odds
post_prob  <- post_odds / (1 + post_odds)

par(mar = c(4, 4, 3, 1))
plot(prev, ppv_v, type = "l", lwd = 2, col = "tomato",
     xlab = "Prior probability (prevalence)",
     ylab = "Posterior probability",
     main = sprintf("LR+ = %.1f: Converting Prior to Posterior", lrp_val),
     ylim = c(0, 1))
lines(prev, post_prob, lwd = 4, lty = 2, col = "steelblue")
abline(0, 1, lty = 3, col = "gray60")
legend("topleft", legend = c("PPV (exact)", "LR approximation", "No update (diagonal)"),
       col = c("tomato", "steelblue", "gray60"),
       lty = c(1, 2, 3), lwd = 2, cex = 0.75)
```


-->

## Belief Updating: A Practical Example
\small
A 72-year-old patient is referred with memory complaints after a GP visit.

\begin{itemize}
\item \textbf{Prior}: based on age, referral context, and brief interview, you estimate a 30\% probability of MCI.
\item \textbf{Test}: a cognitive screening with Sensitivity = 0.85, Specificity = 0.80.
\item \textbf{Result}: the patient scores below the cut-off (positive test).
\end{itemize}

\vfill

$$\text{LR}^+ = \frac{0.85}{1 - 0.80} = 4.25$$

\vfill

Prior odds $= \frac{0.30}{0.70} = 0.43$ $\Rightarrow$ Posterior odds $= 0.43 \times 4.25 = 1.82$

$$\text{Posterior probability} = \frac{1.82}{1 + 1.82} \approx \mathbf{65\%}$$

## Visualising the Update

```{r}
#| echo: FALSE
#| fig.width: 6.5
#| fig.height: 4
#| fig.dpi: 100

prior <- 0.30
se    <- 0.85
sp    <- 0.80
lrp   <- se / (1 - sp)
lrn   <- (1 - se) / sp

# Positive and negative update
prior_o   <- prior / (1 - prior)
post_o_p  <- prior_o * lrp
post_o_n  <- prior_o * lrn
post_p    <- post_o_p / (1 + post_o_p)
post_n    <- post_o_n / (1 + post_o_n)

probs <- c(prior, post_p, post_n)
labs  <- c(sprintf("Prior\n%.0f%%", prior*100),
           sprintf("After\npositive\ntest\n%.0f%%", post_p*100),
           sprintf("After\nnegative\ntest\n%.0f%%", post_n*100))
cols  <- c("gray60", "tomato", "steelblue")

par(mar = c(3, 4, 3, 1))
bp <- barplot(probs, names.arg = labs, col = cols, border = NA,
              ylim = c(0, 1), ylab = "P(condition)",
              main = "Bayesian Update: Prior and Posterior Probabilities")
abline(h = prior, lty = 2, col = "gray40")
text(bp, probs + 0.03, sprintf("%.0f%%", probs * 100), cex = 0.9, font = 2)
```

## What If the Test Is Negative?

Using the same example (prior = 30\%, Se = 0.85, Sp = 0.80):

$$\text{LR}^- = \frac{1 - 0.85}{0.80} = 0.19$$

Prior odds $= 0.43$ $\Rightarrow$ Posterior odds $= 0.43 \times 0.19 = 0.08$

$$\text{Posterior probability} = \frac{0.08}{1 + 0.08} \approx \mathbf{7\%}$$

\vfill
\small
A negative result substantially \emph{reduces} the probability — even though it cannot fully rule out the condition.

## Testing as Progressive Evidence

A single test result is rarely the end of the story.

\vfill

```{=latex}
\begin{tikzpicture}[node distance=0.5cm, auto,
  ev/.style={rectangle, rounded corners, draw=gray!60, fill=gray!10,
             text width=2cm, align=center, minimum height=0.9cm, font=\small}]

\node[ev] (p0) {Clinical interview\\Prior: 30\%};
\node[ev, right=0.8cm of p0] (p1) {Screening test +\\Updated: 65\%};
\node[ev, right=0.8cm of p1] (p2) {Full battery +\\Updated: 90\%};
\node[ev, right=0.8cm of p2] (p3) {Neuroimaging\\Final: 95\%};

\draw[-{Latex}] (p0) -- (p1);
\draw[-{Latex}] (p1) -- (p2);
\draw[-{Latex}] (p2) -- (p3);
\end{tikzpicture}
```

\vfill
\small
Each piece of evidence — test score, clinical interview, anamnesis, imaging — updates the probability. No single source should be interpreted in isolation.

## Integrating All Sources of Evidence

The Bayesian framework makes explicit what good clinical practice already implies:

\begin{itemize}
\item \textbf{Test scores} are one piece of evidence, not a verdict.
\item \textbf{Clinical interview} and \textbf{anamnesis} inform the prior.
\item \textbf{Collateral information} (family, GP referral notes) adds further evidence.
\item Each additional source \emph{updates} the probability estimate.
\end{itemize}

\vfill
\small
A very sensitive test rules \emph{out}; a very specific test rules \emph{in} — but neither replaces clinical judgment.

## Summary

\begin{itemize}
\item Sensitivity and specificity estimate $P(\text{result} \mid \text{condition})$, \textbf{not} $P(\text{condition} \mid \text{result})$.
\item Bayes' theorem is needed to invert this — it requires a prior (pre-test probability).
\item Every test result is an \textbf{update of belief}, not a binary decision.
\item Good clinical assessment integrates test scores with interview, anamnesis, and collateral information as a \textbf{progressive accumulation of evidence}.
\end{itemize}

:::
